Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-09-20 paint-contour-tree.tex
more changes

\newcommand{\bT}{{\bf T}}
\newcommand{\cS}{{\cal S}}
\newcommand{\cB}{{\cal B}}
\newcommand{\cC}{{\cal C}}
\newcommand{\cD}{{\cal D}}
\newcommand{\cE}{{\cal E}}
\newcommand{\cF}{{\cal F}}
\newcommand{\cG}{{\cal G}}
\newcommand{\cH}{{\cal H}}
\newcommand{\cL}{{\cal L}}
\newcommand{\cP}{{\cal P}}
\newcommand{\cV}{{\cal V}}

@@

\subsection{Proving \Lem{light}} \label{sec:light}

@@

We now construct the shrub of unsupported light paths $\cU$. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors

We focus on $\pmax(T)$, and prove the key property of this path decomposition.

\begin{lemma}
\label{lem:geometric}
 Let $p$ be the parent of $q$ in the shrub $\pmax(T)$.  
 Then $|H_q|\leq \frac{3}{5} |H_p|$. 
\end{lemma}
\begin{proof}
 Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
 i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
 As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
 the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
 Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
 Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
 Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

\subsubsection{An asymptotically equivalent weight function}
\label{sec:weight}
Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.

>>>>

\newcommand{\bT}{{\bf T}}
\newcommand{\cS}{{\cal S}}
\newcommand{\cB}{{\cal B}}
\newcommand{\cC}{{\cal C}}
\newcommand{\cD}{{\cal D}}
\newcommand{\cE}{{\cal E}}
\newcommand{\cF}{{\cal F}}
\newcommand{\cG}{{\cal G}}
\newcommand{\cH}{{\cal H}}
\newcommand{\cL}{{\cal L}}
\newcommand{\cP}{{\cal P}}
\newcommand{\cU}{{\cal U}}
\newcommand{\cV}{{\cal V}}

@@

\subsection{The shrub of unsupported paths} \label{sec:light}

@@

We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. But $|H_r| = 0$, implying $p$ is supported.)
Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
roots are supported light paths. \Lem{support} is a direct consequence of the following,
which is the really the main lemma.

\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
\end{lemma}

We proceed with the setup to prove \Lem{cu}.
We focus our attention on some shrub $\cU$.
We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.

\begin{lemma} \label{lem:cu-root} (i) For any $v \in p$ such that $p \in \cU$,
$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
\end{lemma}

\begin{proof} (of \Lem{cu}) We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
\end{proof}

\subsection{The root is everything in $\cU$} \label{sec:weight}

\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
By construction, there is a heavy path (denoted $h(q)$) that is a child of $p$ in $\pmax(T)$,
and an ancestor of $q$. Furthermore, $|H_{h(q)}| > 99|H_q|/100$.
We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.

%
%
%
%We focus on $\pmax(T)$, and prove the key property of this path decomposition.

\begin{lemma}
\label{lem:geometric}
 Let $p$ be the parent of $q$ in the shrub $\cU$. Then $|H_p|\geq \frac{3}{2}|H_q|$. 
\end{lemma}
\begin{proof} We know that $h(q)$ is a child of $p$ in $\pmax(T)$. Consider the base
of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 1$,
$|H_u| \geq 2|H_w| - 1$. Let $b$ be the base of $p$.
We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for small constant $\delta > 0$).
Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| > 99|H_q|/100$,
to complete the proof.

%
%
% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
% Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
for a child $q'$ in $\cU$.

Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
